---
author: David Carslaw
editor_options: 
  markdown: 
    wrap: 72
---

# Wind and Pollution Roses {#sec-windRose}

The wind rose is a very useful way of summarising meteorological data.
It is particularly useful for showing how wind speed and wind direction
conditions vary by year. The `windRose` function can plot wind roses in
a variety of ways: summarising all available wind speed and wind
direction data, plotting individual wind roses by year, and also by
month. The latter is useful for considering how meteorological
conditions vary by season.

Data are summarised by direction, typically by 45 or 30$^\circ$ and by
different wind speed categories. Typically, wind speeds are represented
by different width 'paddles'. The plots show the proportion (here
represented as a percentage) of time that the wind is from a certain
angle and wind speed range.

The `windRose` function also calculates the percentage of 'calms' i.e.
when the wind speed is zero. UK Met Office data assigns these periods to
0 degrees wind direction with valid northerly winds being assigned to
360$^\circ$. The option `calm.thresh` can be used to assign calms to
non-zero wind speeds. For example `calm.thresh = 0.5` will assign wind
speeds *below* 0.5 as calm.

The `windRose` function will also correct for bias when wind directions
are rounded to the nearest 10 degrees but are displayed at angles that
10 degrees is not exactly divisible into e.g. 22.5$^\circ$. When such
data are binned, some angles i.e. N, E, S, W will comprise three
intervals whereas others will comprise two, which can lead to
significant bias. This issue and its solution is discussed by
@droppo2008 and @applequist2012.[^wind-roses-1] [openair]{.pkg} uses a
simple method to correct for the bias by globally rescaling the count in
each wind direction bin by the number of directions it represents
relative to the average. Thus, the primary four directions are each
reduced by a factor of 0.75 and the remaining 12 directions are
multiplied by 1.125.

[^wind-roses-1]: Thanks to Philippe Barneoud of Environment Canada for
    pointing this issue out.

## Example of use

First we load the packages:

```{r}
#| message: false
#| warning: false
library(openair)
library(tidyverse)
```

The function is very simply called as shown for @fig-windroseBasic.

```{r}
#| label: fig-windroseBasic
#| fig-cap: Use of `windRose` function to plot wind speed/direction frequencies. Wind speeds are split into the intervals shown by the scale in each panel. The grey circles show the % frequencies.
#| fig-width: 4
#| fig-height: 5
#| out-width: 60%
windRose(mydata)
```

@fig-windRoseyear highlights some interesting differences between the
years. In 2000, for example, there were numerous occasions when the wind
was from the SSW and 2003 clearly had more occasions when the wind was
easterly. It can also be useful to use `type = "month"` to get an idea
of how wind speed and direction vary seasonally.

```{r}
#| label: fig-windRoseyear
#| fig-cap: Use of `windRose` function to plot wind speed/direction frequencies by year. Wind speeds are split into the intervals shown by the scale in each panel. The grey circles show the 10 and 20% frequencies.
#| fig-width: 8
#| fig-height: 6
windRose(
  mydata,
  type = "year",
  layout = c(4, 2),
  width = 0.5
)
```

The `type` option is very flexible in [openair]{.pkg} and can be used to
quickly consider the dependencies between variables. @sec-cutData
describes the basis of this option in [openair]{.pkg} plot. As an
example, consider the question: what are the meteorological conditions
that control high and low concentrations of PM~10~? By setting
`type = "pm10"`, [openair]{.pkg} will split the PM~10~ concentrations
into four *quantiles* i.e. roughly equal numbers of points in each
level. The plot will then show four different wind roses for each
quantile level, although the default number of levels can be set by the
user --- see `?cutData` for more details. @fig-windRsPM10 shows the
results of setting `type = "pm10"`. For the lowest concentrations of
PM~10~ the wind direction is dominated by northerly winds, and
relatively low wind speeds. By contrast, the highest concentrations
(plot furthest right) are dominated by relatively strong winds from the
south-west. It is therefore very easy to obtain a good idea about the
conditions that tend to lead to high (or low) concentrations of a
pollutant. Furthermore, the `type` option is available in almost all
[openair]{.pkg} functions.

```{r}
#| label: fig-windRsPM10
#| fig-cap: Wind rose for four different levels of PM~10~ concentration. The levels are defined as the four *quantiles* of PM~10~ concentration and the ranges are shown on each of the plot labels.
#| fig-width: 11
#| fig-height: 5
windRose(mydata, type = "pm10", layout = c(4, 1))
```

A comparison of the effect that bias has can be seen by plotting the
following. Note the prominent frequencies for W, E and N in particular
that are due to the bias issue discussed by @applequist2012.

```{r}
#| label: windRoseBias
#| eval: false
## no bias correction
windRose(mydata, angle = 22.5, bias.corr = FALSE)

## bias correction (the default)
windRose(mydata, angle = 22.5)
```

`pollutionRose` is a variant of `windRose` that is useful for
considering pollutant concentrations by wind direction, or more
specifically the percentage time the concentration is in a particular
range. This type of approach can be very informative for air pollutant
species, as demonstrated by Ronald Henry and co-authors in @henry2009.

You can produce similar pollution roses using the `pollutionRose`
function in recent versions of [openair]{.pkg}, e.g. as in
@fig-pollutionRoseBasic:

```{r}
#| label: fig-pollutionRoseBasic
#| fig-cap: NO~x~ pollution rose produced using  `pollutionRose` and default `pollutionRose` settings.
#| fig-width: 5
#| fig-height: 4
#| out-width: 60%
pollutionRose(mydata, pollutant = "nox")
```

`pollutionRose` is *wrapper* for `windRose`. It simply replaces the wind
speed data series in the supplied data set with another variable using
the argument `pollutant` before passing that on to `windRose`. It also
modifies `breaks` to estimate a sensible set of break points for that
pollutant and uses a slightly different set of default options (key to
right, wedge style plot) but otherwise handles arguments just like the
parent `windRose` function.

While @fig-pollutionRoseBasic indicates that higher NO~x~ concentrations
are also associated with the SW, conditioning allows you to be much
informative. For example, conditioning by SO~2~ (@fig-pollRoseNOxO2)
demonstrates that higher NO~x~ concentrations are associated with the SW
and much of the higher SO~2~ concentrations. However, it also highlights
a notable NO~x~ contribution from the E, most apparent at highest SO~2~
concentrations that is obscured in @fig-pollutionRoseBasic by a
relatively high NO~x~ background @fig-pollRoseNOxO2.

```{r}
#| label: fig-pollRoseNOxO2
#| fig-cap: NO~x~ pollution rose conditioned by SO~2~ concentration.
#| fig-width: 10
#| fig-height: 5
pollutionRose(
  mydata,
  pollutant = "nox",
  type = "so2",
  layout = c(4, 1),
  key.position = "bottom"
)
```

`pollutionRose` can also usefully be used to show which wind directions
dominate the overall concentrations. By supplying the option
`statistic = "prop.mean"` (proportion contribution to the mean), a good
idea can be gained as to which wind directions contribute most to
overall concentrations, as well as providing information on the
different concentration levels. A simple plot is shown in
@fig-pollRoseAttr, which clearly shows the dominance of south-westerly
winds controlling the overall mean NO~x~ concentrations at this site.
Indeed, almost half the overall NO~x~ concentration is contributed by
two wind sectors to the south-west. The `polarFreq` function can also
show this sort of information, but the pollution rose is more effective
because both length and colour are used to show the contribution. These
plots are very useful for understanding which wind directions control
the overall mean concentrations.

```{r}
#| label: fig-pollRoseAttr
#| fig-cap: Pollution rose showing which wind directions contribute most to overall mean concentrations.
#| fig-width: 5
#| fig-height: 5
#| out-width: 60%
pollutionRose(
  mydata,
  pollutant = "nox",
  statistic = "prop.mean"
)
```

It is sometimes useful to more clearly understand the contributions from
wind directions that have low frequencies. For example, for a pollution
rose of SO~2~ there are few occurrences of easterly winds making it
difficult to see how the concentration intervals are made up. Try:

```{r}
#| label: pollRoseSO2
#| eval: false
pollutionRose(mydata, pollutant = "so2", seg = 1)
```

However, each wind sector can be normalised to give a probability
between 0 and 1 to help show the variation within each wind sector more
clearly. An example is shown in @fig-pollRoseNorm where for easterly
winds it is now clearer that a greater proportion of the time the
concentration is made up of high SO~2~ concentrations. In this plot each
wind sector is scaled between 0 and 1. Also shown with a black line is
an indication of the wind direction frequency to remind us that winds
from the east occur with a low frequency.

```{r}
#| label: fig-pollRoseNorm
#| fig-cap: SO~2~ pollution rose produced using `pollutionRose` normalised by each wind sector.
#| fig-width: 5
#| fig-height: 4
#| out-width: 60%
pollutionRose(
  mydata,
  pollutant = "so2",
  normalise = TRUE,
  seg = 1,
  cols = "heat"
)
```

## Comparing two meteorological data sets {#sec-comp-met}

The `windRose` function is also useful for comparing two meteorological
data sets. In this case a 'reference' data set is compared with a second
data set. There are many reasons for doing so e.g. to see how one site
compares with another or for meteorological model evaluation (more on
that in later sections). In this case, `ws` and `wd` are considered to
the the reference data sets with which a second set of wind speed and
wind directions are to be compared (`ws2` and `wd2`).

In the next example, data from two data sets from 2024 are imported over
the Internet. The first is hourly data measured data for a site in London and the
second is hourly data modelled data. The data are then joined
together by date to create a single data frame with the two sets of wind
speed and wind direction data. A more realistic example might be to
compare a modelled data set with a reference data set, but the principle
is the same. Such data can be imported using the `{worldmet}` package,
as described in @sec-worldmet.

```{r}
#| label: fig-imprtMet
## example of comparing 2 met sites

# london met data
london_met <- readr::read_rds(
  "https://github.com/openair-project/book/raw/refs/heads/main/assets/data/london_met.rds"
)

# data from model
mod_met <- importAURN("kc1", 2024)

# combine into one data frame and change names - only select ws, wd and date
met_compare <- inner_join(
  select(london_met, date, ws, wd),
  select(mod_met, date, ws, wd),
  suffix = c("_lon", "_yor"),
  by = "date"
)
```

The `windRose` function can then be used to compare the two data sets.
In this case, the plot shown in @fig-windComp shows the difference
between the two data sets with respect to the reference data set (in
this case, the London site). The colours show whether the second data
set (modelled) tends to be positively or negatively biased with respect
to the reference data set (measured). It shows for example:

-  Modelled wind speeds tend to be lower than at measured for wind directions from the south, shown by the blue colouring in @fig-windComp.

-  Wind speeds from the model tend to be slightly higher than measured values when the wind is from the north, shown by the red colouring in @fig-windComp.

- Overall, the modelled wind speeds are 0.67 m s^-1^ lower than the measured values, shown by the text on the plot.

-  There is a bias in the wind direction of +15.9 degrees for the modelled data — shown by both the arrow on the plot and the text.

```{r}
#| label: fig-windComp
#| fig-cap: Wind rose showing the difference between two meteorological data sets. The first (measured) is used to compare against the second (modelled). The colours are used to show whether data tend to be positively or negatively biased with respect to the reference data set. The arrow shows the mean bias in wind direction and the text shows the value of that bias.
#| fig-width: 4.5
#| fig-height: 5.5
#| out-width: 75%
windRose(
  met_compare,
  ws = "ws_lon",
  wd = "wd_lon",
  ws2 = "ws_yor",
  wd2 = "wd_yor"
)
```

Like other [openair]{.pkg} functions, the `type` option can be used to compare the two data sets by other variables, as shown in @fig-windCompSeason for seasonal differences.

```{r}
#| label: fig-windCompSeason
#| fig-cap: Wind rose showing the difference between two meteorological data sets by season. The first (measured) is used to compare against the second (modelled). The colours are used to show whether data tend to be positively or negatively biased with respect to the reference data set. The arrows shows the mean bias in wind direction and the text shows the value of that bias.
#| fig-width: 8
#| fig-height: 9
#| out-width: 90%
windRose(
  met_compare,
  ws = "ws_lon",
  wd = "wd_lon",
  ws2 = "ws_yor",
  wd2 = "wd_yor",
  type = "season"
)
```

