---
author: David Carslaw
---

# Kolmogorov-Zurbenko filters {#sec-ts-filtering}


## Background {#sec-kz-background}

The Kolmogorov-Zurbenko (KZ) filter is an iterative moving-average technique that separates a time series into components operating at distinct temporal scales [@Zurbenko1986; @Yang2010]. Its defining characteristic is simplicity: a plain moving average of window length $m$ is applied $k$ times in succession. This repetition transforms a box-shaped frequency response into one that closely approximates a Gaussian, attenuating high-frequency variability while preserving long-period signals with minimal distortion. The filter is written as KZ($m$, $k$), where $m$ is the half-window length in data steps and $k$ is the number of passes. The effective pass-band boundary (the period below which variation is suppressed) scales roughly as $m\sqrt{k}$, giving the analyst straightforward control over temporal resolution without requiring Fourier transforms or spectral expertise.

The value of the KZ filter in atmospheric and air quality science lies in the decomposition step [@Rao1997; @Eskridge1997]. By applying the filter successively at four increasing window sizes — approximately one day, one week, one month, and one year for hourly data — and taking the *differences* between adjacent filtered series, the raw time series is partitioned into five additive components:

| Component | Typical period | Physical interpretation |
|----|----|----|
| Short | Hours to \~2 days | Diurnal cycles, brief emission pulses |
| Synoptic | \~2–7 days | Frontal passages, weather systems |
| Intermediate | Weeks to \~2 months | Sub-seasonal variability |
| Seasonal | Months to \~1 year | Annual cycle, photochemistry |
| Trend | Multi-year | Long-term emission changes, climate signals |

: Default temporal decomposition produced by `kzFilter()` for hourly data.

These components are mutually exclusive and sum to the original series, making the decomposition physically interpretable and suitable for downstream analysis.

The approach was first demonstrated on ambient ozone data by @RaoZurbenko1994 and @Rao1997, who showed that synoptic-scale meteorological variability can mask long-term ozone trends unless removed. The same logic extends naturally to PM~2.5~, NO~x~, and other regulated pollutants [@Wise2005].

By removing short-term, synoptic and seasonal variability, the KZ decomposition isolates the long-term trend component, which is often the focus of regulatory and epidemiological studies. In this respect, the filters provide a way of doing meteorological normalisation in a straightforward way, which might be useful in many settings --- especially of many years of data are available. The short-term and synoptic components can be analysed separately to understand the influence of local sources and weather patterns on air quality. The intermediate component can reveal sub-seasonal phenomena such as pollution episodes or transport events.

## Kolmogorov-Zurbenko Adaptive filter {#sec-kza-background}

The KZ Adaptive (KZA) filter extends the basic method by detecting structural breaks in the time series and *shrinking* the averaging window near those discontinuities [@Zurbenko1996]. Wherever the algorithm identifies an abrupt change — a step reduction in emissions following a policy intervention, a sensor replacement, or a sudden source change — it uses a shorter window to preserve the sharp feature rather than smearing it across a broad neighbourhood. Away from breaks the filter behaves identically to the standard KZ filter.

The degree of window shrinkage is governed by the `sensitivity` argument: larger values cause more aggressive adaptation. For detecting clear, sudden shifts in air quality (e.g., the introduction of a clean-air zone, COVID-19 lockdown effects, or the commissioning of an industrial scrubber) the KZA filter will generally give a cleaner result than the standard KZ filter.

## Using `kzFilter()` {#sec-kzFilter}

The default window sizes `m = c(25, 169, 721, 8761)` are chosen for hourly data and correspond roughly to one day, one week, one month, and one year. The iteration count `k = 5` provides a near-Gaussian frequency response at each scale. These defaults are appropriate for most ambient monitoring records and rarely need adjustment.

Different window sizes can be used for data at other time resolutions (e.g., 15-minute, daily, 1 Hz data) or to target specific temporal scales. For example, for daily data, windows of 7, 30, 90, and 365 days might be more appropriate. The key is to choose window sizes that capture the desired physical processes while ensuring sufficient data points within each window for stable averaging.

Ten years of data have been uploaded for a background (North Kensington) and roadside site (Marylebone Road) in London for 2016 to 2025, including surface meteorological data. The data can be imported as follows:

```{r}
#| label: kz-import

library(openair)
library(tidyverse)

london_aq <- readr::read_rds("https://github.com/openair-project/book/raw/refs/heads/main/assets/data/london_aq.rds")
```

Next, we calculate the KZ components for a few pollutants at the North Kensington site. The `kzFilter()` function returns a data frame with the original data and new columns for each component.

```{r}
#| label: kz-components
kz_out_pm25 <- kzFilter(
  filter(london_aq, code == "KC1"), 
  pollutant = "pm2.5")

kz_out_o3 <- kzFilter(
  filter(london_aq, code == "KC1"), 
  pollutant = "o3")
```

To show the different components for PM~2.5~, a polar plot can be produced, as shown in @fig-kz-components-pm25. In this case, the short term, synoptic and seasonal variation are considered together. It should be noted that the trend component has been added back to show the overall variation in concentrations. The plot shows that the short-term variation is dominated by low wind speed conditions, typical of local sources at ground level. By contrast, the synoptic and intermediate variation has a much clearer easterly / north-easterly component present also at higher wind speeds, which is dominated by regional transport and the influence of pollution episodes that last a few days (synoptic) to a week or two (intermediate).

```{r}
#| label: fig-kz-components-pm25
#| fig-cap: "Polar plot of the short, synoptic and intermediate components of PM~2.5~ concentrations at the North Kensington site. The trend component has been added back to both components to show the overall variation in concentrations."
#| fig-width: 8
#| fig-height: 5
#| out-width: 90%
kz_out_pm25 |> 
  mutate(
    short = short + trend, 
    synoptic = synoptic + trend, 
    intermediate = intermediate + trend) |> 
polarPlot(
  pollutant = c("short", "synoptic", "intermediate"), 
  limits = c(0, 15), 
  cols = "turbo",
  key.title = "pm25 (ug/m3)",
  key.position = "top"
)
```

For the synoptic component it can be useful to plot a `calendarPlot()` for a particular year. @fig-calendarPlot-synoptic shows the calendar plot for the synoptic component of PM~2.5~ concentrations at the North Kensington site for 2020. The plot shows that the synoptic component is dominated by a few episodes of high concentrations that last a few days and are associated with easterly / north-easterly conditions (as shown in the polar plot above). The plot also shows that there is a lot of variation in the synoptic component, with many days having very low values, and some days having very high values. The trend component has been added back to show the overall variation in concentrations.

```{r}
#| label: fig-calendarPlot-synoptic
#| fig-cap: "Calendar plot of the synoptic component of PM~2.5~ concentrations at the North Kensington site for 2020. The trend component has been added back to show the overall variation in concentrations."
#| fig-width: 8
#| fig-height: 6

mutate(kz_out_pm25, synoptic = synoptic + trend) |> 
  calendarPlot(
    pollutant = "synoptic", 
    year = 2020, 
    cols = colorOpts("batlow", direction = -1),
    windflow = windflowOpts(
      colour = "white"
    ))
```

@fig-kz-seasonal considers the seasonal component of O~3~ concentrations. The plot shows that the seasonal component of O~3~ concentrations is dominated by a strong annual cycle with high concentrations in the spring-summer and low concentrations in the winter. This is typical of O~3~ concentrations in urban areas, where photochemical production dominates in the summer and titration by NO dominates in the winter. Note that the seasonal component shown in @fig-kz-seasonal is smoothly varying because the influence of shorter-term variation has been removed. It can also be useful to combine components e.g. adding the seasonal to the trend (`mutate(kz_out_o3, seasonal = seasonal + trend)`) to show the overall seasonal variation in concentrations on top of a long-term trend.

```{r}
#| label: fig-kz-seasonal
#| fig-cap: "Time plot of the seasonal component of O~3~ concentrations at the North Kensington site."
#| fig-width: 6
#| fig-height: 4

timePlot(kz_out_o3, 
         pollutant = "seasonal", 
         ylab = "o3 (ug/m3)", 
         xlab = "Year"
         )
```

Finally, the `timeVariation()` function can be used to show how the different components vary by hour of the day and month of the year. @fig-kz-timevariation shows the time variation plot for the short-term, synoptic and seasonal components of O~3~ concentrations at the North Kensington site. @fig-kz-timevariation acts as a simple test to see which temporal variations are removed and which remain for different time scales. The short-term component shows a strong diurnal cycle with higher concentrations during the afternoon, and lower concentrations at night. There is no variation in the diurnal plots for the synoptic and seasonal components because shorter-term variation has been removed. The seasonal component shows a strong annual cycle with high concentrations in the spring-summer and low concentrations in the winter, as expected for O~3~ concentrations in urban areas. Similarly, there is no variation in the month of the year plot for the short-term component, and very little for the synoptic. These plots confirm the expected behaviour of the KZ filtering for targetted temporal scales.

```{r}
#| label: fig-kz-timevariation
#| fig-cap: "Time variation plot for the short-term, synoptic and sesasonal components of O~3~ concentrations at the North Kensington site. The short-term component shows a strong diurnal cycle with higher concentrations during the afternoon, and lower concentrations at night. There is no variation in the diurnal plots for the synoptic and seasonal components because shorter-term variation has been removed. The seasonal component shows a strong annual cycle with high concentrations in the spring-summer and low concentrations in the winter, as expected for O~3~ concentrations in urban areas. Similarly, there is no variation in the month of the year plot for the short-term component, and very little for the synoptic."
#| fig-width: 8
#| fig-height: 5
timeVariation(
  kz_out_o3, 
  pollutant = c("short", "synoptic", "seasonal"), 
  ci = FALSE, 
  cols = "Set2")
```

## Using `kzaFilter()` {#sec-kzaFilter}

`kzaFilter()` shares the same interface as `kzFilter()` with one additional argument, `sensitivity`, which controls how aggressively the window is reduced near structural breaks (default 1.0). It is the preferred choice when:

- A known policy change, sensor replacement, or source modification is expected to produce a step change.
- The trend component from `kzFilter()` shows a suspicious sharp feature that may be an artefact of the standard window.
- There is a wish to detect and locate unknown discontinuities.

The choice if `kzFilter()` vs `kzaFilter()` is often a matter of judgement and experimentation. If the KZ trend shows a sharp drop or rise, or rapid variations, that coincides with a known event, it is worth trying the KZA filter to see if it preserves that feature more cleanly. If the KZA trend is much smoother and more interpretable, it may be the better choice for subsequent analysis.

In @fig-kz-kza-compare, the KZ and KZA trends for NO~2~ concentrations at the Marylebone Road site are compared. The KZA filter (red) preserves a sharp drop in NO~2~ in early 2020 that is smoothed out by the standard KZ filter (blue). This drop is related to the COVID-19 lockdown measures that were implemented in the UK in March 2020, which led to a sudden reduction in traffic emissions. The KZA filter is able to capture this abrupt change more effectively than the standard KZ filter, which smooths it out over a longer period.

```{r}
#| label: fig-kz-kza-compare
#| fig-cap: "Comparison of KZ and KZA filters applied to NO~2~ concentrations at the Marylebone Road site. The KZA filter (red) preserves a sharp drop in NO~2~ in early 2020 that is smoothed out by the standard KZ filter (blue)."
#| fig-width: 5
#| fig-height: 4
trend_kz <- kzFilter(
  filter(london_aq, code == "MY1"),
  pollutant = "no2"
) |>
  rename(kz_trend = trend)

trend_kza <- kzaFilter(
  filter(london_aq, code == "MY1"),
  pollutant = "no2",
  sensitivity = 0.8
) |>
  rename(kza_trend = trend)

left_join(trend_kz, select(trend_kza, date, kza_trend), by = "date") |>
  timeAverage(avg.time = "week") |>
  timePlot(
    pollutant = c("kz_trend", "kza_trend"),
    group = TRUE,
    cols = c("steelblue", "tomato"),
    name.pol = c("KZ trend", "KZA trend"),
    ylab = "NO2 trend (ug/m3)",
    linewidth = c(0.5, 0.7)
  )
```

The analysis above for Covid-19 lockdown effects could arguably be better analysed by combining the seasonal and trend components using the KZ filter because the lockdown effects are not a perfect step change and and some of the effects lasted a few weeks to a few months i.e. they would be expected to also show up in the seasonal cycle. An alternative approach is shown below, but not plotted.[^1]

[^1]: Or try the `kzaFilter` with `sensitivity = 0.5` for more structure to capture more of the variation.

```{r}
#| label: kz-kza-combine
#| eval: false
# standard KZ filter
trend_kz <- kzFilter(
   filter(london_aq, code == "MY1"),
   pollutant = "no2")

# combine seasonal and trend components
trend_kz <- mutate(trend_kz, total = seasonal + trend) 
timePlot(trend_kz, pollutant = "total")
```

