---
author: David Carslaw
---

# Utility functions {#sec-utility}

## Selecting data by date {#sec-selectByDate}

Selecting by date/time in R can be intimidating for new users---and time-consuming for all users. The `selectByDate()` function aims to make this easier by allowing users to select data based on the British way of expressing date i.e. d/m/y. This function should be very useful in circumstances where it is necessary to select only part of a data frame.

First load the packages we need.

```{r}
#| message: false
#| warning: false
library(openair)
library(tidyverse)
```

```{r }
## select all of 1999
data.1999 <- selectByDate(mydata, start = "1/1/1999", end = "31/12/1999")
head(data.1999)
tail(data.1999)

## easier way
data.1999 <- selectByDate(mydata, year = 1999)

## more complex use: select weekdays between the hours of 7 am to 7 pm
sub.data <- selectByDate(mydata, day = "weekday", hour = 7:19)

## select weekends between the hours of 7 am to 7 pm in winter (Dec, Jan, Feb)
sub.data <- selectByDate(
  mydata,
  day = "weekend",
  hour = 7:19,
  month = c(12, 1, 2)
)
```

The function can be used directly in other functions. For example, to make a polar plot using year 2000 data:

```{r}
#| eval: false
polarPlot(selectByDate(mydata, year = 2000), pollutant = "so2")
```

## Making intervals --- cutData {#sec-cutData}

The `cutData()` function is a utility function that is called by most other functions but is useful in its own right. Its main use is to partition data in many ways, many of which are built-in to [openair]{.pkg}

Note that all the date-based types e.g. month/year are derived from a column `date`. If a user already has a column with a name of one of the date-based types it will not be used.

For example, to cut data into seasons:

```{r}
#| label: cutdataexam
newdata <- cutData(mydata, type = "season")
head(newdata)
```

This adds a new column `season` that is split into four seasons. There is an option `hemisphere` that can be used to use southern hemisphere seasons when set as `hemisphere = "southern"`.

The `type` can also be another column in a data frame. If this column is numeric, it is divided into `n.levels` *quantiles* (defaulting to four), which is to say bins of roughly equal numbers of PM~10~ concentrations.

```{r}
#| label: cutdata2
newdata <- cutData(mydata, type = "pm10", n.levels = 5)
head(newdata)
```

This use of `cutData()` is quite destructive --- the PM concentraitons are no longer in the data frame. You can use either the `names` or `suffix` arguments to get around this.

```{r}
#| label: cutdata3

# set names manually
newdata <- cutData(mydata, type = "pm10", names = "pollution_bins")
dplyr::glimpse(newdata)

# set a 'suffix' to avoid any columns that would be overlapped
newdata <- cutData(mydata, type = c("o3", "pm10", "so2"), suffix = "_cuts")
dplyr::glimpse(newdata)
```

Most of the time users do not have to call `cutData()` directly because most functions have a `type` option that is used to call `cutData()` directly, e.g.,

```{r}
#| eval: false
polarPlot(mydata, pollutant = "so2", type = "season")
```

However, it can be useful to call `cutData()` *before* supplying the data to a function in a few cases. Commonly, this is to have more control over how the data is conditioned - e.g., to use the `hemisphere` or `n.levels` arguments of `cutData()` directly. More details can be found in the `cutData()` help file.

## Selecting run lengths of values above a threshold --- pollution episodes {#sec-selectRunning}

A seemingly easy thing to do that has relevance to air pollution episodes is to select run lengths of contiguous values of a pollutant above a certain threshold. For example, one might be interested in selecting O~3~ concentrations where there are at least 8 consecutive hours above 90\~ppb. In other words, a selection that combines both a threshold and *persistence*. These periods can be very important from a health perspective and it can be useful to study the conditions under which they occur. But how do you select such periods easily? The `selectRunning` utility function has been written to do this. It could be useful for all sorts of situations e.g.

- Selecting hours when primary pollutant concentrations are persistently high --- and then applying other [openair]{.pkg} functions to analyse the data in more depth.

- In the study of particle suspension or deposition etc. it might be useful to select hours when wind speeds remain high or rainfall persists for several hours to see how these conditions affect particle concentrations.

- It could be useful in health impact studies to select blocks of data where pollutant concentrations remain above a certain threshold.

As an example we are going to consider O~3~ concentrations at a semi-rural site in south-west London (Teddington). The data can be downloaded as follows:

```{r}
#| label: importTed
ted <- openair::importImperial(site = "td0", year = 2005:2009, meteo = TRUE)
## see how many rows there are
nrow(ted)
```

We are going to contrast two pollution roses of O~3~ concentration. The first shows hours where the criterion is not met, and the second where it is met. The subset of hours is defined by O~3~ concentrations above 90 ppb for periods of at least 8-hours i.e. what might be considered as ozone episode conditions.

```{r}
#| label: episodeSelect
ted <- selectRunning(ted, pollutant = "o3", threshold = 90, run.len = 8)
```

The `selectRunning()` function returns a new column that flags whether the condition is met or not. The user can control the text provided, which by default is "yes" and "no", as well as the name of the column that is appended (defaulting to `"criterion"`).

```{r}
#| label: criterion
table(ted$criterion)
```

Now we are going to produce two pollution roses shown in @fig-RunningRes. Note, however that many other types of analysis could be carried out now the data have been partitioned.

```{r}
#| label: fig-RunningRes
#| fig-cap: Example of using the `selectRunning` function to select episode hours to produce pollution roses of O~3~ concentration.
#| fig-width: 8
#| fig-height: 5
#| out-width: 90%
pollutionRose(
  ted,
  pollutant = "o3",
  type = "criterion"
)
```

The results are shown in @fig-RunningRes. The pollution rose for the "no" criterion (left plot of @fig-RunningRes) shows that the highest O~3~ concentrations tend to occur for wind directions from the south-west, where there is a high proportion of measurements. By contrast, the when the criterion is met (right plot of @fig-RunningRes) is very different. In this case there is a clear set of conditions where these criteria are met i.e. lengths of at least 8-hours where the O~3~ concentration is at least 90 ppb. It is clear the highest concentrations are dominated by south-easterly conditions i.e. corresponding to easterly flow from continental Europe where there has been time to the O~3~ chemistry to take place.

The code below shows (as an example), that the summer of 2006 had a high proportion of conditions where the criterion was met.

```{r}
#| eval: false
timeProp(
  ted,
  pollutant = "o3",
  proportion = "criterion",
  avg.time = "month",
  cols = "viridis"
)
```

It is also useful to consider what controls the highest NO~x~ concentrations at a central London roadside site. For example, the code below (not plotted) shows very strongly that the persistently highest NO~x~ concentrations are dominated by south-westerly winds. As mentioned earlier, there are many other types of analysis that can be carried out now the data set identifies where the criterion is or is not met.

```{r}
#| label: runningNOx
#| eval: false
episode <- selectRunning(
  mydata,
  pollutant = "nox",
  threshold = 500,
  run.len = 5
)

pollutionRose(episode, pollutant = "nox", type = "criterion")
```

Note that `selectRunning()` will fail if there are duplicate dates within the data frame. If multiple sites worth of data are contained within a single data frame, use `type = "site"` (or whatever column identifies different sites).

## Calculating rolling means {#sec-rollingMean}

Some air pollution statistics such as for O~3~ and particulate matter are expressed as rolling means and it is useful to be able to calculate these. It can also be useful to help smooth-out data for clearer plotting. The `rollingMean()` function makes these calculations. One detail that can be important is that for some statistics a mean is only considered valid if there are a sufficient number of valid readings over the averaging period. Often there is a requirement for at least 75% data capture. For example, with an averaging period of 8 hours and a data capture threshold of 75%, at least 6 hours are required to calculate the mean.

The function is called as follows; in this case to calculate 8-hour rolling mean concentrations of O~3~.

```{r}
#| label: selectBydate
mydata <- rollingMean(
  mydata,
  pollutant = "o3",
  width = 8,
  new.name = "rollingo3",
  data.thresh = 75
)
tail(mydata)
```

Note that calculating rolling means shortens the length of the data set. In the case of O~3~, no calculations are made for the last 7 hours.

Like `selectRunning()`, `rollingMean()` will fail if there are duplicate dates within the data frame; users can again use `type = "site"` or similar to calculate rolling means for different monitoring stations.

Type `?rollingMean` into R for more details.

## Calculating rolling quantiles (percentiles) {#sec-rollingQuantile}

There are many reasons why it is useful to calculate a running quantile for time series data. First, it is useful to flexibly show how high or low quantile trends change over time. Second, a low quantile (such as 1 or 2%) can often be useful for determining a 'background' concentration. For example, for fast reponse measurements (e.g. at a 1 second resolution) such as that from mobile measurements or fixed location measurements, the background value is often identified using a low quantile rolling measurement [@Padilla2021; @Farren2024].

The `rollingQuantile` function provides a fast way of calculating one or more quantiles in a rime series. The function can also take account of a data threshold so that quantiles are not calculated if there are too few data available in a particular time window.

An example is shown below and plotted in @fig-rollingQuant. In this example, we choose a rolling window width of 744 (about a month) and calculate quantile levels at 0.05 and 0.95. @fig-rollingQuant shows the jump in NO~2~ concentrations at the start of 2003 for the high quantile level but not the lower one. As mentioned elsewhere, this increase in NO~2~ was related to an increase in primary NO~2~ emissions related to bus aftertreatment technology.

```{r}
#| label: fig-rollingQuant
#| fig-cap: Example of using the `rollingQuantile` function to calculate low (0.05) and high (0.95) rolling quantiles.
#| fig-width: 6
#| fig-height: 4
data(mydata) # reload the original data

# calculate quantiles
quantile_data <- rollingQuantile(
  mydata,
  pollutant = "no2",
  width = 744,
  data.thresh = 75,
  probs = c(0.05, 0.95)
)

# look at the data
dplyr::glimpse(quantile_data)

# plot it
timePlot(
  quantile_data,
  pollutant = c("q_no2_0.05", "q_no2_0.95"),
  group = TRUE,
  linetype = 1,
  cols = c("tomato", "turquoise4"),
  ylab = "no2 (ug/m3)"
)
```

## Gaussian smoothing {#sec-gauss-smooth}

There are numerous ways of smoothing and aggregating time series (and other data). One common and powerful technique is *Gaussian smoothing* where a Gaussian window slides along the data for each time step and smooths the data. The basic idea is that the smoothing process puts more weight on data close to the time in question and less further away with the weighting being determined by the σ value in the Gaussian equation.

Gaussian smoothing is powerful and flexible and has many advantages over simpler methods such as a moving average. Gaussian smoothing results in a smoother curve, and better preserves trends, and reduced edge artefacts. 

In the example shown in @fig-gauss-smooth, the data have been plotted as weekly averages to make the raw data easier to see. The smoothed line provides a clear indication of how NO~2~ concentrations have varied over many years. In this case `sigma` was set to 744, which corresponds to about one month.

It should be noted `GassianSmooth` will work with any time series data e.g. at 1 Hz so should be useful for processing noisy fast response data.

```{r}
#| label: fig-gauss-smooth
#| fig-cap: Example of using the `GassianSmooth` function to smooth NO~2~ concentrations with sigma set at about 1 month.
#| fig-width: 6
#| fig-height: 4
data(mydata) # reload the original data

gauss_smooth <- GaussianSmooth(
  mydata,
  pollutant = "no2",
  sigma = 744
)

# show the data
dplyr::glimpse(gauss_smooth)

# plot weekly means of orginal data and smoothed
timePlot(
  timeAverage(gauss_smooth, avg.time = "week"),
  pollutant = c("no2", "smooth_no2"),
  group = TRUE,
  linetype = 1,
  cols = c("tomato", "turquoise4"),
  linewidth = c(1, 3),
  ylab = "no2 (ug/m3)"
)
```

## Whittaker-Eilers smoothing {#sec-whittaker}

### Background

The Whittaker-Eilers smoother [@eilers2003] is a powerful and flexible method for smoothing and interpolating time series data. The method is based on penalised least squares: it seeks a smooth series that is close to the observed data while penalising roughness (i.e., large differences between adjacent values). The balance between fidelity to the data and smoothness is controlled by a single parameter `lambda` — larger values produce smoother results.

Key advantages of the Whittaker-Eilers approach over simpler methods such as a moving average include:

- It handles missing values (`NA`) naturally, interpolating across gaps without requiring imputation.
- It is computationally efficient even for long time series [@eilers2003].
- The order of the roughness penalty `d` can be varied: `d = 2` (the default) penalises curvature, `d = 1` penalises slope and effectively performs linear interpolation across gaps.
- The `lambda` parameter gives direct, intuitive control over the degree of smoothing.

### Standard smoothing

The simplest use of `WhittakerSmooth()` is to produce a smoothed version of a pollutant time series. In the example below, NO~2~ concentrations are smoothed with `lambda = 2000`. The function adds a new column `smooth_no2` to the data frame.

Where the aim is to reduce noise, setting `lambda = NA` will use Generalised Cross Validation (GCV) to choose an optimal `lambda` value automatically, though this can be computationally expensive for large datasets. However, for smoothing purposes, it is often more useful to experiment with a range of `lambda` values to find one that produces a result that is consistent with the aims of smoothing data in the first instance.

```{r}
#| label: whittaker-smooth
mydata_smooth <- WhittakerSmooth(
  mydata,
  pollutant = "no2",
  lambda = 2000
)
dplyr::glimpse(mydata_smooth)
```

@fig-whittaker-smooth shows the first 744 hours (approximately one month) of raw and smoothed NO~2~ data. The Whittaker smoother closely tracks the data while removing short-term noise, making the underlying variation much easier to interpret.

```{r}
#| label: fig-whittaker-smooth
#| fig-cap: "Whittaker-Eilers smoothed NO~2~ concentrations (red) plotted against the raw hourly data (blue) for the first 744 hours of `mydata`. Lambda was set to 20."
#| fig-width: 6
#| fig-height: 4
mydata_smooth |>
  slice_head(n = 744) |>
  timePlot(
    pollutant = c("smooth_no2", "no2"),
    group = TRUE,
    linetype = 1,
    linewidth = c(2, 0.3),
    cols = c("tomato", "steelblue"),
    name.pol = c("Whittaker smooth", "Raw NO2"),
    ylab = "NO2 (µg/m³)"
  )
```

The `lambda` value can be increased substantially for longer time series spanning many years. Because smoothing a multi-year hourly record requires much greater regularisation, values in the range `lambda = 1e6` to `lambda = 1e10` are often appropriate. If `lambda = NA`, Generalised Cross Validation (GCV) is used to choose `lambda` automatically, though this can be computationally expensive for large datasets.

A major advantage of the Whittaker-Eilers approach is that a user can confidently "set and forget" their `lambda` value. Because `lambda` acts as an absolute penalty, the local smoothing characteristics will remain identical regardless of how much new data is appended to the series; as a time series grows, the Whittaker smooth will always deliver a predictable, consistent outcome. In contrast, methods like LOESS (locally-weighted regression) or GAMs often rely on a relative span or fraction of the total data. To maintain that same visually and mathematically consistent smoothing effect as a dataset expands, users of those methods must constantly recalculate and shrink their smoothing parameters.

### Baseline estimation

Beyond smoothing, `WhittakerSmooth()` can be used for *baseline estimation* via Asymmetric Least Squares (ALS) by supplying the `p` argument. When `p` is set to a small value (e.g., 0.001), the smoother is asymmetrically penalised: it is strongly discouraged from rising above the data but not from falling below it. This forces the smooth curve to track the *lower envelope* of the time series — the background or baseline — rather than the central tendency.

The function returns two new columns:

- `<pollutant>_baseline`: the estimated background concentration.
- `<pollutant>_increment`: the difference between the observed concentration and the baseline, representing the local enhancement above background.

This decomposition is particularly useful in air pollution studies, where it is common to want to separate a *background* component (e.g., regional or photochemical NO~2~) from local *increments* (e.g., fresh traffic emissions). The increment can then be used as the basis for further analysis such as polar plots or wind roses to identify source directions.

### Comparison with rolling low quantiles

A common alternative approach for estimating a background is to use a rolling low quantile (e.g., the 1st percentile over a sliding window). @fig-whittaker-baseline compares the Whittaker baseline with a 1st-percentile rolling quantile computed over a 50-hour window.

```{r}
#| label: fig-whittaker-baseline
#| fig-cap: "Comparison of the Whittaker baseline (blue) and a 50-hour rolling 1st-percentile (red) against the raw NO~2~ data (green) for the first 1000 hours of `mydata`. The Whittaker baseline varies smoothly and continuously, while the rolling quantile produces a stepped series."
#| fig-width: 6
#| fig-height: 4
WhittakerSmooth(mydata, pollutant = "no2", lambda = 2000, p = 0.001) |>
  rollingQuantile(pollutant = "no2", width = 50, probs = 0.01) |>
  slice_head(n = 1000) |>
  timePlot(
    pollutant = c("no2_baseline", "q_no2_0.01", "no2"),
    group = TRUE,
    linewidth = c(2, 2, 0.4),
    ylab = "NO2 (µg/m3)",
    cols = "tol.bright",
    name.pol = c("Whittaker", "Rolling Quantile", "Raw NO2 data")
  )
```

As @fig-whittaker-baseline illustrates, the rolling quantile produces a stepped, discontinuous baseline that can change abruptly as observations enter or leave the rolling window. By contrast, the Whittaker baseline varies smoothly and continuously, which is more physically plausible for a background concentration and produces better-behaved increments for downstream analysis. The Whittaker approach also handles missing data naturally, whereas gaps can distort the effective window length for rolling statistics.

## Kolmogorov-Zurbenko filters {#sec-kz-filter}

### Background {#sec-kz-background}

The Kolmogorov-Zurbenko (KZ) filter is an iterative moving-average technique that separates a time series into components operating at distinct temporal scales [@Zurbenko1986; @Yang2010]. Its defining characteristic is simplicity: a plain moving average of window length $m$ is applied $k$ times in succession. This repetition transforms a box-shaped frequency response into one that closely approximates a Gaussian, attenuating high-frequency variability while preserving long-period signals with minimal distortion. The filter is written as KZ($m$, $k$), where $m$ is the half-window length in data steps and $k$ is the number of passes. The effective pass-band boundary (the period below which variation is suppressed) scales roughly as $m\sqrt{k}$, giving the analyst straightforward control over temporal resolution without requiring Fourier transforms or spectral expertise.

The value of the KZ filter in atmospheric and air quality science lies in the decomposition step [@Rao1997; @Eskridge1997]. By applying the filter successively at four increasing window sizes — approximately one day, one week, one month, and one year for hourly data — and taking the *differences* between adjacent filtered series, the raw time series is partitioned into five additive components:

| Component | Typical period | Physical interpretation |
|----|----|----|
| Short | Hours to \~2 days | Diurnal cycles, brief emission pulses |
| Synoptic | \~2–7 days | Frontal passages, weather systems |
| Intermediate | Weeks to \~2 months | Sub-seasonal variability |
| Seasonal | Months to \~1 year | Annual cycle, photochemistry |
| Trend | Multi-year | Long-term emission changes, climate signals |

: Default temporal decomposition produced by `kzFilter()` for hourly data.

These components are mutually exclusive and sum to the original series, making the decomposition physically interpretable and suitable for downstream analysis.

The approach was first demonstrated on ambient ozone data by @RaoZurbenko1994 and @Rao1997, who showed that synoptic-scale meteorological variability can mask long-term ozone trends unless removed. The same logic extends naturally to PM~2.5~, NO~x~, and other regulated pollutants [@Wise2005].

By removing short-term, synoptic and seasonal variability, the KZ decomposition isolates the long-term trend component, which is often the focus of regulatory and epidemiological studies. In this respect, the filters provide a way of doing meteorological normalisation in a straightforward way, which might be useful in many settings --- especially of many years of data are available. The short-term and synoptic components can be analysed separately to understand the influence of local sources and weather patterns on air quality. The intermediate component can reveal sub-seasonal phenomena such as pollution episodes or transport events. 

### Kolmogorov-Zurbenko Adaptive filter {#sec-kza-background}

The KZ Adaptive (KZA) filter extends the basic method by detecting structural breaks in the time series and *shrinking* the averaging window near those discontinuities [@Zurbenko1996]. Wherever the algorithm identifies an abrupt change — a step reduction in emissions following a policy intervention, a sensor replacement, or a sudden source change — it uses a shorter window to preserve the sharp feature rather than smearing it across a broad neighbourhood. Away from breaks the filter behaves identically to the standard KZ filter.

The degree of window shrinkage is governed by the `sensitivity` argument: larger values cause more aggressive adaptation. For detecting clear, sudden shifts in air quality (e.g., the introduction of a clean-air zone, COVID-19 lockdown effects, or the commissioning of an industrial scrubber) the KZA filter will generally give a cleaner result than the standard KZ filter.

### Using `kzFilter()` {#sec-kzFilter}

The default window sizes `m = c(25, 169, 721, 8761)` are chosen for hourly data and correspond roughly to one day, one week, one month, and one year. The iteration count `k = 5` provides a near-Gaussian frequency response at each scale. These defaults are appropriate for most ambient monitoring records and rarely need adjustment. 

Different window sizes can be used for data at other time resolutions (e.g., 15-minute, daily, 1 Hz data) or to target specific temporal scales. For example, for daily data, windows of 7, 30, 90, and 365 days might be more appropriate. The key is to choose window sizes that capture the desired physical processes while ensuring sufficient data points within each window for stable averaging.

Ten years of data have been uploaded for a background (North Kensington) and roadside site (Marylebone Road) in London for 2016 to 2025, including surface meteorological data. The data can be imported as follows:

```{r}
#| label: kz-import
london_aq <- readr::read_rds("https://github.com/openair-project/book/raw/refs/heads/main/assets/data/london_aq.rds")
```

Next, we calculate the KZ components for a few pollutants at the North Kensington site. The `kzFilter()` function returns a data frame with the original data and new columns for each component.

```{r}
#| label: kz-components
kz_out_pm25 <- kzFilter(
  filter(london_aq, code == "KC1"), 
  pollutant = "pm2.5")

kz_out_o3 <- kzFilter(
  filter(london_aq, code == "KC1"), 
  pollutant = "o3")
```

To show the different components for PM~2.5~, a polar plot can be produced, as shown in @fig-kz-components-pm25. In this case, the short term and synoptic variation are considered together. It should be noted that the trend component has been added back to both the short and synoptic components to show the overall variation in concentrations. The plot shows that the short-term variation is dominated by low wind speed conditions, typical of local sources at ground level. By contrast, the synoptic variation has a much clearer easterly / north-easterly component present also at higher wind speeds, which is dominated by regional transport and the indluence of pollution episodes that last a few days.

```{r}
#| label: fig-kz-components-pm25
#| fig-cap: "Polar plot of the short and synoptic components of PM~2.5~ concentrations at the North Kensington site. The trend component has been added back to both components to show the overall variation in concentrations."
#| fig-width: 6
#| fig-height: 4
kz_out_pm25 |> 
  mutate(short = short + trend, synoptic = synoptic + trend) |> 
polarPlot(
  pollutant = c("short", "synoptic"), 
  limits = c(0, 15), 
  cols = "turbo",
  key.title = "pm25\n(ug/m3)"
)
```

For the synoptic component it can be useful to plot a `calendarPlot()` for a particular year. @fig-calendarPlot-synoptic shows the calendar plot for the synoptic component of PM~2.5~ concentrations at the North Kensington site for 2020. The plot shows that the synoptic component is dominated by a few episodes of high concentrations that last a few days and are associated with easterly / north-easterly conditions (as shown in the polar plot above). The plot also shows that there is a lot of variation in the synoptic component, with many days having very low values, and some days having very high values. The trend component has been added back to show the overall variation in concentrations.

```{r}
#| label: fig-calendarPlot-synoptic
#| fig-cap: "Calendar plot of the synoptic component of PM~2.5~ concentrations at the North Kensington site for 2020. The trend component has been added back to show the overall variation in concentrations."
#| fig-width: 8
#| fig-height: 6

mutate(kz_out_pm25, synoptic = synoptic + trend) |> 
  calendarPlot(
    pollutant = "synoptic", 
    year = 2020, 
    cols = colorOpts("batlow", direction = -1),
    windflow = TRUE)
```

 @fig-kz-seasonal considers the seasonal component of O~3~ concentrations. The plot shows that the seasonal component of O~3~ concentrations is dominated by a strong annual cycle with high concentrations in the spring-summer and low concentrations in the winter. This is typical of O~3~ concentrations in urban areas, where photochemical production dominates in the summer and titration by NO dominates in the winter. Note that the seasonal component shown in @fig-kz-seasonal is smoothly varying because the influence of shorter-term variation has been removed. It can also be useful to combine components e.g. adding the seasonal to the trend (`mutate(kz_out_o3, seasonal = seasonal + trend)`) to show the overall seasonal variation in concentrations on top of a long-term trend.

```{r}
#| label: fig-kz-seasonal
#| fig-cap: "Time plot of the seasonal component of O~3~ concentrations at the North Kensington site."
#| fig-width: 6
#| fig-height: 4

timePlot(kz_out_o3, 
         pollutant = "seasonal", 
         ylab = "o3 (ug/m3)", 
         xlab = "Year"
         )
```

Finally, the `timeVariation()` function can be used to show how the different components vary by hour of the day and month of the year. @fig-kz-timevariation shows the time variation plot for the short-term, synoptic and seasonal components of O~3~ concentrations at the North Kensington site. @fig-kz-timevariation acts as a simple test to see which temporal variations are removed and which remain for different time scales. The short-term component shows a strong diurnal cycle with higher concentrations during the afternoon, and lower concentrations at night. There is no variation in the diurnal plots for the synoptic and seasonal components because shorter-term variation has been removed. The seasonal component shows a strong annual cycle with high concentrations in the spring-summer and low concentrations in the winter, as expected for O~3~ concentrations in urban areas. Similarly, there is no variation in the month of the year plot for the short-term component, and very little for the synoptic. These plots confirm the expected behaviour of the KZ filtering for targetted temporal scales.

```{r}
#| label: fig-kz-timevariation
#| fig-cap: "Time variation plot for the short-term, synoptic and sesasonal components of O~3~ concentrations at the North Kensington site. The short-term component shows a strong diurnal cycle with higher concentrations during the afternoon, and lower concentrations at night. There is no variation in the diurnal plots for the synoptic and seasonal components because shorter-term variation has been removed. The seasonal component shows a strong annual cycle with high concentrations in the spring-summer and low concentrations in the winter, as expected for O~3~ concentrations in urban areas. Similarly, there is no variation in the month of the year plot for the short-term component, and very little for the synoptic."
#| fig-width: 8
#| fig-height: 5
timeVariation(
  kz_out_o3, 
  pollutant = c("short", "synoptic", "seasonal"), 
  ci = FALSE, 
  cols = "Set2")
```

### Using `kzaFilter()` {#sec-kzaFilter}

`kzaFilter()` shares the same interface as `kzFilter()` with one additional argument, `sensitivity`, which controls how aggressively the window is reduced near structural breaks (default 1.0). It is the preferred choice when:

- A known policy change, sensor replacement, or source modification is expected to produce a step change.
- The trend component from `kzFilter()` shows a suspicious sharp feature that may be an artefact of the standard window.
- There is a wish to detect and locate unknown discontinuities.

The choice if `kzFilter()` vs `kzaFilter()` is often a matter of judgement and experimentation. If the KZ trend shows a sharp drop or rise, or rapid variations, that coincides with a known event, it is worth trying the KZA filter to see if it preserves that feature more cleanly. If the KZA trend is much smoother and more interpretable, it may be the better choice for subsequent analysis.

In @fig-kz-kza-compare, the KZ and KZA trends for NO~2~ concentrations at the Marylebone Road site are compared. The KZA filter (red) preserves a sharp drop in NO~2~ in early 2020 that is smoothed out by the standard KZ filter (blue). This drop is related to the COVID-19 lockdown measures that were implemented in the UK in March 2020, which led to a sudden reduction in traffic emissions. The KZA filter is able to capture this abrupt change more effectively than the standard KZ filter, which smooths it out over a longer period.

```{r}
#| label: fig-kz-kza-compare
#| fig-cap: "Comparison of KZ and KZA filters applied to NO~2~ concentrations at the Marylebone Road site. The KZA filter (red) preserves a sharp drop in NO~2~ in early 2020 that is smoothed out by the standard KZ filter (blue)."
#| fig-width: 5
#| fig-height: 4
trend_kz <- kzFilter(
  filter(london_aq, code == "MY1"),
  pollutant = "no2"
) |>
  rename(kz_trend = trend)

trend_kza <- kzaFilter(
  filter(london_aq, code == "MY1"),
  pollutant = "no2",
  sensitivity = 0.8
) |>
  rename(kza_trend = trend)

left_join(trend_kz, select(trend_kza, date, kza_trend), by = "date") |>
  timeAverage(avg.time = "week") |>
  timePlot(
    pollutant = c("kz_trend", "kza_trend"),
    group = TRUE,
    cols = c("steelblue", "tomato"),
    name.pol = c("KZ trend", "KZA trend"),
    ylab = "NO2 trend (ug/m3)",
    linewidth = c(0.5, 1)
  )
```


## Aggregating data by different time intervals {#sec-timeAverage}

Aggregating data by different averaging periods is a common and important task. There are many reasons for aggregating data in this way:

- Data sets may have different averaging periods and there is a need to combine them. For example, the task of combining an hourly air quality data set with a 15-minute average meteorological data set. The need here would be to aggregate the 15-minute data to 1-hour before merging.

- It is extremely useful to consider data with different averaging times straightforwardly. Plotting a very long time series of hourly or higher resolution data can hide the main features and it would be useful to apply a specific (but flexible) averaging period to the data for plotting.

- Those who make measurements during field campaigns (particularly for academic research) may have many instruments with a range of different time resolutions. It can be useful to re-calculate time series with a common averaging period; or maybe help reduce noise.

- It is useful to calculate statistics other than means when aggregating e.g. percentile values, maximums etc.

- For statistical analysis there can be short-term autocorrelation present. Being able to choose a longer averaging period is sometimes a useful strategy for minimising autocorrelation.

In aggregating data in this way, there are a couple of other issues that can be useful to deal with at the same time. First, the calculation of proper vector-averaged wind direction is essential. Second, sometimes it is useful to set a minimum number of data points that must be present before the averaging is done. For example, in calculating monthly averages, it may be unwise to not account for data capture if some months only have a few valid points.

When a data capture threshold is set through `data.thresh` it is necessary for `timeAverage()` to know what the original time interval of the input time series is. The function will try and calculate this interval based on the most common time gap (and will print the assumed time gap to the screen). This works fine most of the time but there are occasions where it may not e.g. when very few data exist in a data frame. In this case the user can explicitly specify the interval through `interval` in the same format as `avg.time` e.g. `interval = "month"`. It may also be useful to set `start.date` and `end.date` if the time series do not span the entire period of interest. For example, if a time series ended in October and annual means are required, setting `end.date` to the end of the year will ensure that the whole period is covered and that `data.thresh` is correctly calculated. The same also goes for a time series that starts later in the year where `start.date` should be set to the beginning of the year.

All these issues are (hopefully) dealt with by the `timeAverage()` function. The options are shown below, but as ever it is best to check the help that comes with the [openair]{.pkg} package.

To calculate daily means from hourly (or higher resolution) data:

```{r}
#| label: dailyAve
daily <- timeAverage(mydata, avg.time = "day")
daily
```

Monthly 95th percentile values:

```{r}
#| label: percentile95
monthly <- timeAverage(
  mydata,
  avg.time = "month",
  statistic = "percentile",
  percentile = 95
)
monthly
```

2-week averages but only calculate if at least 75% of the data are available:

```{r}
#| label: twoweek
twoweek <- timeAverage(mydata, avg.time = "2 week", data.thresh = 75)
twoweek
```

Note that `timeAverage()` has a `type` option to allow for the splitting of variables by a grouping variable. The most common use for `type` is when data are available for different sites and the averaging needs to be done on a per site basis. That being said, unlike other data utility functions, `timeAverage()` will *warn* if there are duplicate dates in the input data frame but still allow the calculation to continue; occasionally averaging across multiple sites may be a legitimate thing to do (e.g., averaging across many low-cost air quality sensors to calculate a more accurate 'consensus' time series).

First, retaining by site averages:

```{r}
#| label: timeAvType
# import some data for two sites
dat <- importUKAQ(c("kc1", "my1"), year = 2011:2013)

# annual averages by site
timeAverage(dat, avg.time = "year", type = "site")
```

Retain site name and site code:

```{r}
#| label: timeAvTypeCode
# can also retain site code
timeAverage(dat, avg.time = "year", type = c("site", "code"))
```

Average all data across sites (drops site and code).

```{r}
#| label: timeAvNoType
timeAverage(dat, avg.time = "year")
```

`timeAverage()` also works the other way in that it can be used to derive higher temporal resolution data e.g. hourly from daily data or 15-minute from hourly data. An example of usage would be the combining of daily mean particle data with hourly meteorological data. There are two ways these two data sets can be combined: either average the meteorological data to daily means or calculate hourly means from the particle data. The `timeAverage()` function when used to 'expand' data in this way will repeat the original values the number of times required to fill the new time scale. In the example below we calculate 15-minute data from hourly data. As it can be seen, the first line is repeated four times and so on.

```{r}
#| label: timeAvgFill
data15 <- timeAverage(mydata, avg.time = "15 min", fill = TRUE)
head(data15, 20)
```

The `timePlot()` function can apply this function directly to make it very easy to plot data with different averaging times and statistics.

## Calculating percentiles {#sec-calcPercentile}

`calcPercentile()` makes it straightforward to calculate percentiles for a single pollutant. It can take account of different averaging periods, data capture thresholds --- see @sec-timeAverage for more details.

`calcPercentile()` shares many of its arguments with `timeAverage()`, such as `type`, `data.thresh`, `start.date`, and `end.date`. It also has the argument `prefix` to define how the columns it creates are labelled.

For example, to calculate the 25, 50, 75 and 95th percentiles of O~3~ concentration by year:

```{r}
#| label: calcpercentile
calcPercentile(
  mydata,
  pollutant = "o3",
  percentile = c(25, 50, 75, 95),
  avg.time = "year"
)
```

## Correlation matrices {#sec-corPlot}

Understanding how different variables are related to one another is always important. However, it can be difficult to easily develop an understanding of the relationships when many different variables are present. One of the useful techniques used is to plot a *correlation matrix*, which provides the correlation between all pairs of data. The basic idea of a correlation matrix has been extended to help visualise relationships between variables by [@friendly2002] and [@Sarkar2008].

The `corPlot()` function shows the correlation coded in three ways: by shape (ellipses), colour and the numeric value. The ellipses can be thought of as visual representations of scatter plot. With a perfect positive correlation a line at 45 degrees positive slope is drawn. For zero correlation the shape becomes a circle --- imagine a 'fuzz' of points with no relationship between them.

With many variables it can be difficult to see relationships between variables i.e. which variables tend to behave most like one another. For this reason hierarchical clustering is applied to the correlation matrices to group variables that are most similar to one another (if `cluster = TRUE`.)

An example of the `corPlot()` function is shown in @fig-corPlot. In this Figure it can be seen the highest correlation coefficient is between PM~10~ and PM~2.5~ (r = 0. 84) and that the correlations between SO~2~, NO~2~ and NO~x~ are also high. O~3~ has a negative correlation with most pollutants, which is expected due to the reaction between NO and O~3~. It is not that apparent in @fig-corPlot that the order the variables appear is due to their similarity with one another, through hierarchical cluster analysis. In this case we have chosen to also plot a *dendrogram* that appears on the right of the plot. Dendrograms provide additional information to help with visualising how groups of variables are related to one another. Note that dendrograms can only be plotted for `type = "default"` i.e. for a single panel plot.

```{r}
#| label: fig-corPlot
#| fig-cap: Example of a correlation matrix showing the relationships between variables.
#| fig-width: 7
#| fig-height: 6
#| out-width: 75%
corPlot(mydata, dendrogram = TRUE)
```

Note also that the `corPlot()` accepts a `type` option, so it possible to condition the data in many flexible ways, although this may become difficult to visualise with too many panels. For example:

```{r}
#| eval: false
corPlot(mydata, type = "season")
```

When there are a very large number of variables present, the `corPlot()` function is a very effective way of quickly gaining an idea of how variables are related. As an example (not plotted) it is useful to consider the hydrocarbons measured at Marylebone Road. There is a lot of information in the hydrocarbon plot (about 40 species), but due to the hierarchical clustering it is possible to see that isoprene, ethane and propane behave differently to most of the other hydrocarbons. This is because they have different (non-vehicle exhaust) origins. Ethane and propane results from natural gas leakage whereas isoprene is biogenic in origin (although some is from vehicle exhaust too). It is also worth considering how the relationships change between the species over the years as hydrocarbon emissions are increasingly controlled, or maybe the difference between summer and winter blends of fuels and so on.

```{r}
#| eval: false
hc <- importUKAQ(site = "my1", year = 2005, hc = TRUE)
## now it is possible to see the hydrocarbons that behave most
## similarly to one another
corPlot(hc)
```