---
author: Jack Davison
execute: 
  warning: false
  error: false
  message: false
---

# Customising openair plots {#sec-annotating-openair}

A frequently asked question about [openair]{.pkg} and a commonly requested feature is how to annotate plots. Since all [openair]{.pkg} functions now return [ggplot2]{.pkg} objects, users have considerably more flexibility in adding annotations using the extensive set of tools available in [ggplot2]{.pkg} and its extensions. Embedding annotation options directly inside every [openair]{.pkg} function would make them unnecessarily complex and still not cover every customisation users may want. Instead, because plots are returned as standard [ggplot2]{.pkg} objects, users can build annotations incrementally, adding layers as needed.

There are several types of annotation that can be useful, including text, shapes, lines, and shading. Because many [openair]{.pkg} plots consist of multiple facets, it is also important to understand how to target specific panels when adding annotations. The approaches described here apply broadly to all [openair]{.pkg} plots - the only differences concern the coordinate system used.

The foundation of annotation in the new [openair]{.pkg} is the [ggplot2]{.pkg} layering system. Annotations can be added with standard geoms such as `ggplot2::geom_text()`, `ggplot2::geom_label()`, `ggplot2::geom_rect()`, `ggplot2::geom_segment()`, or using helper functions like ``ggplot2::annotate()`. Since the output is a normal ggplot object, users can modify it by adding layers with the `+` operator, adjusting scales, or adding specialised annotation layers from packages such as [ggrepel]{.pkg}, [ggforce]{.pkg}, or [patchwork]{.pkg} for more complex layouts.

One quirk of [openair]{.pkg} to be aware of is that plotting functions do not return a *plot*, they return an *openair object* which contains the plot, the plot data, and the function call. One tip is to use `plot = FALSE`, which tells the function to not print the plot at time of object creation.

```{r}
#| label: create-scatter-object
#| output: false
library(openair)

scatter_object <- scatterPlot(
  mydata,
  x = "no2",
  y = "nox",
  avg.time = "month",
  plot = FALSE
)

# extract data
scatter_object$data

# extract plot
scatter_object$plot
```

 To make life easier, lets load the [ggplot2]{.pkg} package:

```{r}
#| label: load-ggplot2
#| output: false
library(ggplot2)
```

A good introduction to tweaking plots is using `labs()` to change the axis labels. With `ggplot2` you simply add (with `+`) new layers. Feel free to use `quickText()` directly here too for automatic formatting.

```{r}
#| label: fig-scatter-plot-labs
#| fig-cap: "Using `labs()` to change axis labels and title."
scatter_object$plot +
  labs(
    x = quickText("NO2 (ug/m3)"),
    y = quickText("NOx (ug/m3)"),
    title = "Scatter Plot",
    subtitle = quickText("Monthly NOx roughly correlates with NO2")
  )
```

## Annotations

### Adding text

To add text, use `ggplot2::annotate()` with `geom = "text"` and specify the coordinates, colour and label. In this example, the text 'This is high!' is added to the right of the plot at a y-value of 300.

```{r}
#| label: fig-scatter-plot-annotate-text
#| fig-cap: "Adding text annotations with `annotate()`."
scatter_object$plot +
  annotate(
    geom = "text",
    x = 48.5,
    y = 300,
    colour = "blue",
    label = "This is high!"
  )
```

### Adding a shaded area

For a shaded area, use `geom = "rect"` and specify the coordinates of the rectangle. In this example, a semi-transparent orange rectangle is added to highlight a region of interest. Note that the `alpha` parameter controls the transparency of the rectangle, with values closer to 0 being more transparent and values closer to 1 being more opaque. We also use `fill` over `colour` to specify the fill colour of the rectangle; `colour` could be used to specify the border colour if desired.

```{r}
#| label: fig-scatter-plot-annotate-rect
#| fig-cap: "Adding a shaded area with `annotate()`."
scatter_object$plot +
  annotate(
    geom = "rect",
    xmin = 40,
    xmax = 50,
    ymin = 100,
    ymax = 200,
    fill = "orange",
    alpha = 0.5
  ) +
  annotate(
    geom = "text",
    x = 45,
    y = 150,
    color = "orange4",
    fontface = "bold",
    label = "Look at this!"
  )
```

### Adding an arrow

To add an arrow, use `geom = "segment"` and specify the coordinates of the start and end points, along with the `arrow` parameter to define the arrowhead. In this example, an arrow is added pointing from (40, 150) to (60, 250). Note that `arrow(ends = "both")` creates arrowheads at both ends of the segment, and `linewidth` controls the thickness of the arrow line. The `arrow()` function also has additional parameters such as `angle` and `length` to further customize the appearance of the arrowheads.

```{r}
#| label: fig-scatter-plot-annotate-arrow
#| fig-cap: "Adding an arrow with `annotate()`."
scatter_object$plot +
  annotate(
    geom = "path",
    x = c(40, 60),
    y = c(150, 250),
    linewidth = 1,
    arrow = arrow(ends = "both")
  )
```

### Highlight a specific point

To highlight a point, you can use `geom = "point"` and specify the coordinates of the point you want to highlight. For example, to highlight the point with the maximum NO~2~ concentration, you can first extract that point from the data and then add it as an annotation. Note that different [openair]{.pkg} plots will have different `data` structures, so explore what's available for your specific plot of interest.

```{r}
#| label: fig-scatter-plot-annotate-point
#| fig-cap: "Highlighting a specific point with `annotate()`."

# can extract a specific point from the scatterplot data
max_no2_point <-
  scatter_object$data[scatter_object$data$no2 == max(scatter_object$data$no2), ]

# add this as an annotation
scatter_object$plot +
  annotate(
    geom = "point",
    x = max_no2_point$no2,
    y = max_no2_point$nox,
    size = 5,
    shape = 1
  )

```

### Add a reference line

To add a reference line, you use `ggplot2::geom_hline()` or `ggplot2::geom_vline()`. In this example, a horizontal dashed purple line is added at y = 200, along with an annotation to label it as "Limit". The `linetype` parameter controls the line type (e.g., dashed, dotted), and `color` specifies the color of the line.

In this case, we set `x = I(0.01)`. The `I()` function tells `ggplot2` to ignore the data scale (e.g., 30 to 70 units of NO2) and instead use a relative scale (0 at the far left, 1 at the far right). This is really helpful - for example, you may create a lot of different `timePlot()` plots with different x-axis scales - `x = I(0.01)` will *always* mean "nudged a little bit from the far left of the plot".

```{r}
#| label: fig-scatter-plot-annotate-hline
#| fig-cap: "Adding a reference line with `geom_hline()` and annotating it."
scatter_object$plot +
  geom_hline(
    yintercept = 200,
    linetype = 2,
    color = "purple"
  ) +
  annotate(
    geom = "text",
    y = 205,
    x = I(0.01),
    vjust = 0,
    hjust = 0,
    label = "Limit",
    color = "purple"
  )
```

### Working with dates

Working with dates (e.g., for plots like `timePlot()`) is a step more challenging, but is mostly identical to what has already been covered. `ISOdate()` is convenient way to create date(times) in R to plug into `annotate()` and similar functions. In this example, we add a vertical dashed line at a specific date (June 1, 2002) along with an annotation to label. The `xintercept` parameter specifies the position of the vertical line, and the `annotate()` function is used to add text at a specific date and y-value. The `hjust` and `vjust` parameters control the horizontal and vertical justification of the text, respectively.

```{r}
#| label: fig-timeplot-annotate-date
#| fig-cap: "Adding a vertical line at a specific date with `geom_vline()` and annotating it."
no2_timeplot <- timePlot(mydata, "no2", avg.time = "month", plot = FALSE)$plot

no2_timeplot +
  geom_vline(
    xintercept = ISOdate(2002, 6, 1, 0),
    linetype = 2,
    color = "grey25"
  ) +
  annotate(
    geom = "text",
    x = ISOdate(2002, 5, 5, 0),
    y = I(0.975),
    label = "What happened\nhere?",
    hjust = 1,
    vjust = 1,
    color = "grey25"
  )
```

### Working with radial axes

For "radial" plots (e.g., `polarPlot()`) the same principles apply, but the coordinates are different. You simply need to think in "radial" coordinates - the x-axis is the 'compass directions' in degrees (0-360) and the y-axis is the radial axis progressing out from the centre. In this example, a vertical dashed line is added at 225 degrees, along with two rectangles to highlight specific angular regions, and text annotations to label these regions as "A", "B", and "C". The `annotate()` function is used to add both the rectangles and the text, with appropriate coordinates for each annotation.

```{r}
#| label: fig-polarplot-annotate
#| fig-cap: "Adding annotations to a polar plot, using radial coordinates."
no2_polarplot <- polarPlot(mydata, "no2", plot = FALSE)$plot

no2_polarplot +
  geom_vline(
    xintercept = 225,
    linetype = 2,
    linewidth = 1
  ) +
  annotate(
    geom = "rect",
    xmin = 60,
    xmax = 150,
    ymin = 5,
    ymax = 20,
    fill = NA,
    color = "black",
    linewidth = 1
  ) +
  annotate(
    geom = "rect",
    xmin = -20,
    xmax = 20,
    ymin = 8,
    ymax = 18,
    fill = NA,
    color = "black",
    linewidth = 1
  ) +
  annotate(
    geom = "text",
    x = 0,
    y = 15,
    label = "A",
    size = 8
  ) +
  annotate(
    geom = "text",
    x = 105,
    y = 15,
    label = "B",
    size = 8
  ) +
  annotate(
    geom = "label",
    x = 225,
    y = 15,
    label = "C",
    size = 8,
    linewidth = 1
  )
```

### Multi-panel plots

When `type` is used in an `openair` plot, multiple panels ("facets" in [ggplot2]{.pkg} speak) are created. Further, some functions like `calendarPlot()` are automatically faceted. You may wish to annotate only a single panel. Use the `layout` argument to specify the panel of interest. 

Below, we highlight Christmas day on a `calendarPlot()`. Note that `layout = 12` means "the twelfth panel, progressing from left-to-right, top-to-bottom". This argument could take a vector - e.g., `layout = c(5, 12)` - which would annotate the fifth and twelfth panels. 

```{r}
#| label: fig-calendarplot-annotate
#| fig-cap: "Annotating a specific panel in a multi-panel plot using the `layout` argument."
multipanel <- calendarPlot(mydata, "no2", year = 2000, plot = F)$plot

multipanel +
  annotate(
    geom = "rect",
    xmin = 2.5,
    xmax = 3.5,
    ymin = 1.5,
    ymax = 2.5,
    layout = 12,
    color = "black",
    fill = NA
  )
```

### Assembly plots

A few `openair` plots are actually multiple plots in one. You can typically tell this because the different "panels" are arranged on an irregular grid (e.g., `timeVariation()`) or are comprised of very different plot types (e.g., `conditionalEval()`). These plots are powered by the [patchwork]{.pkg} package, which makes it simple to extract and modify sub-panels. For `timeVariation()`, the "top" panel can be accessed with `[[1]]` and the "bottom" three with `[[2]][[1]]`, `[[2]][[2]]`, and `[[2]][[3]]`.^[Note that it may easier to use `variationPlot()` and use [patchwork]{.pkg} directly yourself, depending on your use-case.]

```{r}
#| label: fig-timevariation-annotate
#| fig-cap: "Annotating a specific panel in an assembly plot, using `timeVariation()` as an example."
timevar_no2 <- timeVariation(mydata, "no2", plot = FALSE)

timevar_no2$main.plot[[2]][[1]] <- timevar_no2$main.plot[[2]][[1]] +
  geom_hline(yintercept = 50)

plot(timevar_no2)
```

## Facets

One of the sets of arguments passed through `...` could be described as "facet" arguments, as these influence the appearance of 'faceted' (i.e., multi-panel) plots. All of these arguments get passed straight through to either `ggplot2::facet_wrap()` or `ggplot2::facet_grid()` - see the documentation for these functions to learn more.

To demonstrate, lets consider this heatmap plot:

```{r}
#| label: fig-heatmap-1
#| fig-cap: "A standard heatmap plot."
#| fig-height: 3.5
#| fig-width: 8
heatmap_data <- selectByDate(mydata, year = 2003:2005)
trendLevel(heatmap_data, type = "year")
```

In this case, all three of the panels share the same x- and y-scales. Despite this, some readers may prefer that each of the axes are labelled, even if they are identical. The `axes` argument (which defaults to `"margins"`) controls this; it can either be `"all_x"`, `"all_y"` or `"all"` to make axes appear.

```{r}
#| label: fig-heatmap-2
#| fig-cap: "A heatmap with all axes labelled."
#| fig-height: 3.5
#| fig-width: 8
trendLevel(heatmap_data, type = "year", axes = "all")
```

@fig-heatmap-2 clearly shows that our data finishes at the end of June 2005, meaning the rest of that panel is empty. It may be appropriate to simply exclude the rest of 2005 from our plot. To do so, we can use the `scales` argument, which can take `"free"`, `"free_y"`, `"free_x"` or `"fixed"` (the default).

```{r}
#| label: fig-heatmap-3
#| fig-cap: "A heatmap with a 'free' x-axis."
#| fig-height: 3.5
#| fig-width: 8
trendLevel(heatmap_data, type = "year", scales = "free_x")
```

@fig-heatmap-3 is an improvement, but the way in which the 2005 panel is "stretched" is aesthetically unappealing. The `space` argument allows the width or heights of panels to match the scale of data displayed on the axes. When only one `type` is specified, this argument can either take `"fixed"` (the default), `"free_x"` or `"free_y"`, and will force the plot to be 1 dimensional row or column of plots rather than wrapped into a square grid. In this case, our plot is already a row of three so no change is apparent. If there are two values given to `type`, and therefore the plot is a 2D matrix of panels, `space` can also take `"free"` to vary both x and y.

```{r}
#| label: fig-heatmap-4
#| fig-cap: "A heatmap with a 'free' x-axis and panel widths that match the data displayed on the axes."
#| fig-height: 3.5
#| fig-width: 8
trendLevel(heatmap_data, type = "year", scales = "free_x", space = "free_x")
```

All of `nrow`, `ncol`, `scales`, `space`, `axes`, `axis.labels`, `switch` and `strip.position` can be passed through `...` to be used to customise multi-panel plot layout and appearance. 

Note that not all of these arguments will always work in all functions. For example, plots on radial axes like `polarPlot()` fix their x-axis (radial axis) range to be from 0 to 360 to represent compass directions, so `scales = "free_x"` will not do anything.

::: {.callout-note}
### Legacy Options

Users of older versions of [openair]{.pkg} may recall the `x.relation`, `y.relation` and `layout` options. These mapped onto [lattice]{.pkg} arguments, which pre-dated [ggplot2]{.pkg} as the plotting engine for [openair]{.pkg}. These arguments are automatically remapped onto the modern ones, so old code should still work, but it is recommended to use the new arguments directly. A conversion table is provided below:

```{r}
#| label: legacy-map
#| echo: false

gt_nrow <-
  tibble::tibble(
  legacy_layout = c("`c(2, 5)`"),
  new_ncol = c("`2`"),
  new_nrow = c("`5`")
) |>
  gt::gt() |>
  gt::tab_spanner(
    label = "legacy",
    columns = legacy_layout
  ) |>
  gt::tab_spanner(
    label = "new",
    columns = c(new_ncol, new_nrow)
  ) |>
  gt::cols_label(
    legacy_layout = gt::md("`layout`"),
    new_ncol = gt::md("`ncol`"),
    new_nrow = gt::md("`nrow`")
  ) |>
  gt::fmt_markdown(gt::everything())

gt_layout <-
  tibble::tibble(
    legacy_x.relation = c("`'free'`", "`'same'`", "`'free'`", "`'same'`"),
    legacy_y.relation = c("`'free'`", "`'free'`", "`'same'`", "`'same'`"),
    new_scales = c("`'free'`", "`'free_y'`", "`'free_x'`", "`'fixed'`")
  ) |>
  gt::gt() |>
  gt::tab_spanner(
    label = "legacy",
    columns = c(legacy_x.relation, legacy_y.relation)
  ) |>
  gt::tab_spanner(
    label = "new",
    columns = new_scales
  ) |>
  gt::cols_label(
    legacy_x.relation = gt::md("`x.relation`"),
    legacy_y.relation = gt::md("`y.relation`"),
    new_scales = gt::md("`scales`")
  ) |>
  gt::fmt_markdown(gt::everything())

gt::gt_group(
  gt_nrow,
  gt_layout
)
```

:::

## Themes

With [ggplot2]{.pkg} it is simple to control the "theme" of your plot (i.e., all non-data elements) using the `ggplot2::theme()` function. There is an excellent blog post about [ggplot2]{.pkg} stlying at <https://tidyverse.org/blog/2025/10/ggplot2-styling/>, but below we demonstrate a few examples of tweaking the plot style.

```{r}
#| label: fig-scatter-plot-theme
#| fig-cap: "Customising the plot theme with `theme()`."
scatter_object$plot +
  theme(
    text = element_text(size = 15),
    panel.grid.minor = element_line(
      linetype = 3,
      color = "cadetblue2",
      linewidth = 0.5
    ),
    panel.grid.major = element_line(
      linetype = 2,
      color = "cadetblue",
      linewidth = 0.5
    ),
    axis.title.y = element_text(
      angle = 0,
      vjust = 0.5
    ),
    axis.text.x = element_text(
      angle = 90,
      vjust = 0.5
    ),
    panel.background = element_rect(
      fill = "grey95"
    ),
    axis.line = element_line(
      linewidth = 3,
      lineend = "square"
    )
  )
```

It is also possible to set a global 'complete' theme. This could be a default [ggplot2]{.pkg} theme, or one from an extension package. This could be helpful to ensure your [openair]{.pkg} plots align with other plots you're already making in [ggplot2]{.pkg}.

```{r}
#| label: fig-timeplot-theme
#| fig-cap: "Setting a global theme to control the styling of all openair plots."
theme_set(theme_classic())
timePlot(mydata, avg.time = "week")
```

To reset this behaviour, reset the global theme to `theme_grey()`. This is the [ggplot2]{.pkg} default, and makes all of the [openair]{.pkg} functions restore their default behaviour. If you'd like to actually use `theme_grey()`, set your global theme to something like `theme_grey(10.01)` which is different *enough* to disable [openair]{.pkg}'s overrides.

```{r}
#| label: reset-theme
theme_set(theme_gray())
```

Note that some functions do their own `theme()` manipulations, so your mileage may vary as to which of your options are honoured - or whether they do unexpected things to your plot!

A related function is `guides()`, which can be useful for controlling how axes and legends behave. These are typically set within `openair` functions, but can be overridden by more advanced users. Useful arguments include `position` and `angle`. For categorical scales, `check.overlap` or `n.dodge` can help mitigate overlapping axis labels.

```{r}
#| label: fig-scatter-plot-guides
#| fig-cap: "Customising axes with `guides()`."
scatter_object$plot +
  guides(
    x = guide_axis(
      position = "top",
      angle = 90,
      minor.ticks = TRUE
    )
  )
```

## Extensions

There are [many `ggplot2` extensions](https://exts.ggplot2.tidyverse.org/) out there which you can take advantage of in extending your [openair]{.pkg} plots.

One useful example is [patchwork]{.pkg}, which provides an elegant API for combining multiple plots together into one assembly. This is how `timeVariaiton()` works under-the-hood, but users should feel free to assemble plots however they see fit.

The package documentation for [patchwork]{.pkg} is excellent and can be found [here](https://patchwork.data-imaginist.com/).

```{r}
#| label: fig-patchwork-assembly
#| fig-cap: "Using the patchwork package to combine multiple openair plots into one assembly. The `tag` argument is being used to label each one with a letter "
library(patchwork)

p_varplot_hour <- variationPlot(
  mydata,
  "nox",
  "hour",
  caption = "",
  plot = FALSE,
  tag = "A"
)$plot

p_varplot_weekday <- variationPlot(
  mydata,
  "nox",
  "weekday",
  caption = "",
  plot = FALSE,
  tag = "B"
)$plot

p_timeplot <- timePlot(
  mydata,
  "nox",
  avg.time = "month",
  plot = FALSE,
  tag = "C"
)$plot

p_heatmap <- trendLevel(mydata, type = "default", plot = FALSE, tag = "D")$plot

(((p_varplot_hour | p_varplot_weekday) / p_timeplot) | p_heatmap) +
  plot_layout(widths = c(0.7, 0.3)) &
  theme(plot.margin = unit(rep(0, 4), "cm"))
```

Another really useful package is [geomtextpath]{.pkg}, which combines labels and lines and paths for nicer annotations:

```{r}
#| label: geomtextpath-annotation
#| fig-cap: "Using the geomtextpath package to add text along a path, and horizontal reference lines with labels."
library(geomtextpath)

scatter_object$plot +
  annotate(
    geom = "textpath",
    x = c(40, 60),
    y = c(100, 300),
    label = "Things go up!",
    arrow = arrow()
  ) +
  geom_texthline(
    yintercept = 250,
    color = "blue",
    label = "limit value",
    hjust = 0.1,
    linetype = 2
  )
```

An interesting package is [legendry]{.pkg}, which has a lot of different functions for tweaking legends and guides. Of interest may be `legendry::guide_colring()`, which wraps a continuous legend into a doughnut. This could be useful for cyclical variables like wind direction.

This example highlights that sometimes you may need to take on further tweaks as you get deeper into customising your plots. Here, it was necessary to reset the colour scale to set appropriate labels and breaks.

```{r}
#| label: fig-legendry-guide
#| fig-cap: "Using the legendry package to create a custom legend for a scatter plot."
scatterPlot(mydata, z = "wd", plot = FALSE)$plot +
  guides(
    color = legendry::guide_colring()
  ) +
  scale_color_gradientn(
    colours = openair::openColours("hue"),
    breaks = seq(0, 360, 90),
    labels = c("N", "E", "S", "W", "N")
  )
```

Many extension packages usefully provide themes that can be easily set globally to apply to [openair]{.pkg} plots. For example, for those working with or for the UK Government, the [afcharts]{.pkg} package exports `afcharts::theme_af()` for accessible, Government Statistical Service-approved plot styling.

```{r}
#| label: fig-afcharts-theme
#| fig-cap: "Setting a global theme from an extension package, in this case `afcharts::theme_af()`, and creating a `timePlot()`."
theme_set(afcharts::theme_af())

timePlot(
  mydata,
  c("nox", "no2"),
  avg.time = "month",
  group = TRUE,
  cols = "gaf.cat"
)
```

```{r}
#| label: fig-windrose-theme
#| fig-cap: "Customising the theme of a plot from an extension package, in this case `afcharts::theme_af()`."
#| fig-height: 6
windRose(mydata, cols = "gaf.seq")
```

```{r}
#| label: reset-theme-again
theme_set(theme_gray())
```

