---
author: 
- David Carslaw
- Jack Davison
abstract: This section outlines the different ways users can access the abundant air quality data in the UK. The main functions provide easy access to hourly data and other statistical summaries such as annual means and data capture rates. Easy access is also provided to site meta data such as longitude, latitude, site type and details of the pollutants measured.
aliases:
  - sections/maps/maps-network.html
---

# Accessing UK Air Quality Data {#sec-importUKAQ}

## Accessing data

The UK has a surprisingly large amount of air quality data that is publicly accessible. The main UK AURN archive and regional (England, Scotland, Wales and Northern Ireland) together with Imperial College London's London Air Quality Network (LAQN) are important and large databases of information that allow free public access. Storing and managing data in this way has many advantages including consistent data format, and underlying high quality methods to process and store the data.

[openair]{.pkg} has a core function that provides users with extensive access to UK air quality data; `importUKAQ()`. [WSP](https://www.wsp.com/en-gb/) (formerly Ricardo) has provided `.RData` files (R workspaces) for several important air quality networks in the UK. These files are updated on a daily basis. This approach requires a link to the Internet to work. The work of Trevor Davies at WSP is greatly appreciated in making all the data available. The networks available through `importUKAQ()` include:

-   The UK national network, the [Automatic Urban and Rural Network](https://uk-air.defra.gov.uk/networks/network-info?view=aurn). This is the main UK network.

-   The UK "devolved" networks:

    -   [Air Quality Scotland](http://www.scottishairquality.scot/) network.

    -   [Air Quality Wales](https://airquality.gov.wales/) network.

    -   [Air Quality England](https://www.airqualityengland.co.uk/) network of sites.

    -   [Northern Ireland](https://www.airqualityni.co.uk) network of sites.

-   Locally managed AQ networks in England. These are sites operated in most cases by Local Authorities but may also include monitoring from other programmes, industry and airports. The location and purpose of these sites differs from the national network which is governed by strict rules of the air quality directives. As a result there is a broad range of site types, equipment and data quality practices. For more information see [here](https://uk-air.defra.gov.uk/networks/network-info?view=nondefraaqmon). These data represent information from about 15 different local air quality networks.

Also available is `importImperial()` for accessing data from the sites operated by Imperial College London[^uk-air-quality-data-1], primarily including the [The London Air Quality Network](https://www.londonair.org.uk/LondonAir/Default.aspx).

[^uk-air-quality-data-1]: The data were first accessible when the
    Environmental Research Group was based at King's College London.

::: {.callout-tip}
## What about Europe?

When analysing UK air quality data, it may be useful to compare it to data from the UK's neighbours in Europe. Historically, the [saqgetr]{.pkg} allowed for access to European monitoring data, and [openair]{.pkg} provided the `importEurope()` function for a simplified interface for the same data source. Unfortunately, as of February 2024, this service is no longer supported.

The European Environment Agency (EEA) now provides an air quality download service (<https://eeadmz1-downloads-webapp.azurewebsites.net/>), which includes an API. The `openair` family now includes the [euroaq]{.pkg} package, which allows you to easily access this through R - see <https://openair-project.github.io/euroaq/> for more information.
:::

Many users download hourly data from [GOV UK](https://get-air-pollution-data.service.gov.uk/). This is a useful facility but does have some limitations and frustrations, many of which are overcome using `openair` to download the data:

1. It is quick to select a range of sites, pollutants and periods, and it is possible to download several long time series in one go.

1. Storing the data as .RData objects is very efficient as they are about four times smaller than .csv files (which are already small). This means the data downloads quickly and saves bandwidth.

1. The function completely avoids any need for data manipulation or setting time formats, time zones etc.

Some examples of usage are shown below. First, lets load the packages we need.

```{r}
#| warning: false
#| message: false
#| cache: false
library(openair)
library(tidyverse)
```

## Site Meta Data

### National networks

The first question is, what sites are available and what do they measure? Users can access the details of air pollution monitoring sites using the `importMeta()` function. The user only needs to provide the network name and (optionally) whether all data should be returned and whether certain periods should be considered. By default only site type, latitude and longitude are returned.

```{r}
#| label: importMeta
#| warning: false
#| message: false
aurn_meta <- importMeta(source = "aurn")
aurn_meta
```

Or return much more detailed data that includes which pollutants are measured at each site and site start / end dates. The option `all = TRUE` should be added.

```{r}
#| label: importMetaAll
#| eval: true
aurn_meta <- importMeta(source = "aurn", all = TRUE)

# what comes back?
glimpse(aurn_meta)
```

Note that `importMeta()` can import information for several networks at once e.g. `source = c("aurn", "saqn")`. Further, a convenience value for `source` is `"ukaq"`, which returns metadata for all networks accessible via `importUKAQ()` (AURN, AQE, SAQN, WAQN, NI, and local networks).

Often it is useful to consider sites that were open in a particular year or were open for a duration of years. This can be done using the `year` argument. When `year` is a range such as `year = 2010:2020`, only sites that were open across that range of years will be returned. This option is especially useful for trend analysis when there might be an interest in extracting only sites that were measuring over the period of interest. Furthermore, if `all = TRUE` is used, supplying a year (or years) will select only specific pollutants that were measured during the period of interest.[^uk-air-quality-data-2]

[^uk-air-quality-data-2]: Note that while a site may have been in
    operation for a long time, it is not necessarily the case that all
    pollutants were measured since the start of the site.

For example, to check the number of sites that were open from 2010 to 2022 in the AURN and SAQN combined:

```{r}
sites_2010_2022 <- importMeta(
  source = c("aurn", "saqn"),
  year = 2010:2022
)

nrow(sites_2010_2022)
```

`importMeta()` also supports filtering by pollutant directly, without needing to filter the returned data frame manually. The `pollutant` argument accepts one or more pollutant names (case-insensitive), and `"hc"` can be used as a shorthand for all hydrocarbons. For example, to find all Urban Traffic sites on the AURN that measure NO~2~:

```{r}
#| warning: false
#| message: false
no2_sites <- importMeta(
  source = "aurn",
  all = TRUE,
  pollutant = "no2",
  site_type = "Urban Traffic"
)

nrow(no2_sites)
```

Additional filtering arguments include `site` (pattern-matched site name), `code` (exact site code), and `duplicate` (set `TRUE` to retain sites that appear in multiple networks rather than de-duplicating them).

::: callout-tip
#### Use `importMeta()` as a way to select sites to import

One of the most useful aspects of `importMeta()` is to use it as a basis to identify site codes to then import data. For example, to import data from the AURN for sites that have been in operation from 2005 to 2020:

```{r}
#| eval: false
sites_2005_2020 <- importMeta(
  source = "aurn",
  year = 2005:2020
)

all_aq_data <- importUKAQ(
  site = sites_2005_2020$code,
  source = sites_2005_2020$source,
  year = 2005:2020
)
```
:::

### Finding sites near a location

`importMeta()` can also find sites near a specific location by supplying `lat` and `lng` coordinates. The returned data will include a `distance_km` column and be sorted by distance. This is useful when assessing pollution exposure at a receptor (e.g., a school or hospital) or when looking for monitors near a potential pollution source (e.g., an industrial installation or quarry).

The `max_dist` argument limits results to sites within a given radius (in km), and `max_n` caps the number of sites returned. These can be combined to find, say, the four nearest sites within 5 km. [openairmaps]{.pkg} provides the `convertPostcode()` function as a convenient way to obtain coordinates from a UK postcode. The below example finds the four closest monitoring stations within 5km of Great Ormond Street Hospital, a famous children's hospital in London.

```{r}
#| eval: true
hospital_coords <- list(
  lat = 51.52221,
  lng = -0.119898
)

importMeta(
  source = c("aurn", "aqe"),
  year = 2022,
  lat = hospital_coords$lat,
  lng = hospital_coords$lng,
  max_dist = 2,
  max_n = 4
)
```

### Plot Sites on a Map

To easily visualise entire monitoring networks, consider using the [openairmaps]{.pkg} R package. This package can be installed from CRAN, similar to [openair]{.pkg}.

```{r}
#| eval: false
install.packages("openairmaps")
```

This package contains the `networkMap()` function which acts as a wrapper around `importMeta()` and returns a detailed map, with many options for customisation. For example, sites can be clustered together to avoid clutter, and an optional "control menu" can be added to filter for certain sites (e.g., different site types, shown below). All of the filtering arguments described above --- `year`, `pollutant`, `site_type`, `lat`, `lng`, `max_dist`, `max_n` --- can be passed directly to `networkMap()` and are forwarded on to `importMeta()`.

```{r}
#| label: fig-openairmapsNetworkMap
#| cache: false
#| fig-cap: Plotting some English Air Quality stations using the `openairmaps` package.
library(openairmaps)
networkMap(source = c("aurn", "aqe"), control = "site_type")
```

When spatial filtering is applied, `networkMap()` clearly shows the 'target' coordinate as a white marker and, if relevant, the `max_dist` radius as a circle.

```{r}
#| label: fig-openairmapsNetworkMap2
#| cache: false
#| fig-cap: A map of AURN and AQE monitoring stations local to Great Ormond Street.
networkMap(
  source = c("aurn", "aqe"),
  year = 2022,
  lat = hospital_coords$lat,
  lng = hospital_coords$lng,
  max_dist = 2
)
```

As well as providing all of the functionality of `importMeta()`, `networkMap()` provides several additional arguments for customisation:

- `control`: Any column of the equivalent `importMeta()` dataset, which is used to create a "layer control" menu to allow readers to filter for certain sites. The `control` option is quite opinionated, and selects an appropriate style of layer control depending on the column selected (e.g., pollutants are switched between, whereas multiple site types can be selected at once).

- `cluster`: By default, if greater than 25 markers are plotted then they default to be clustered together until users zoom in close. This default behaviour improves the appearance and performance of the HTML map widget. The `cluster` argument allows you to turn this feature on and off.

- `provider`: Any number of the [leaflet]{.pkg} providers (see `leaflet::providers`). The default for this is a typical 'street map' and a satellite map, both of which can be toggled between.

## Monitoring Data

### Hourly data

To import data, you can use the `importUKAQ()` function. Some examples are below.

```{r}
#| eval: false

# import all pollutants from Marylebone Rd from 2000:2005
mary <- importUKAQ(site = "my1", year = 2000:2005)

# import nox, no2, o3 from Marylebone Road and Nottingham Centre for 2000
thedata <- importUKAQ(
  site = c("my1", "nott"),
  year = 2000,
  pollutant = c("nox", "no2", "o3")
)

# import over 30 years of Mace Head O3 data!
o3 <- importUKAQ(site = "mh", year = 1987:2019)

# import hydrocarbon data from Marylebone Road
hc <- importUKAQ(site = "my1", year = 2008, hc = TRUE)

# Import data from the AQE network (York data in this case)
yk13 <- importUKAQ(site = "yk13", year = 2018, source = "aqe")

# Import data from the AURN *and* AQE network! Different networks are
# automatically detected
duo <- importUKAQ(site = c("my1", "yk13"), year = 2020)

# Or, `source` can be specified - useful for ambiguous codes (some 'locally
# managed' sites may have overlapping site codes w/ nationally managed
# sites, e.g., "AD1")
ad1_saqn <- importUKAQ(site = "ad1", year = 2020, source = "saqn")
ad1_eng <- importUKAQ(site = "ad1", year = 2020, source = "local")
```

And to include basic meta data when importing air pollution data:

```{r}
#| label: importWithMeta
## default metadata
kc1 <- importUKAQ(site = "kc1", year = 2018, meta = TRUE)
glimpse(kc1)

## custom metadata - anything in `importMeta()`
kc1_zagglom <- importUKAQ(
  site = "kc1",
  year = 2018,
  meta = TRUE,
  meta_columns = c("zone", "agglomeration")
)
glimpse(kc1_zagglom)
```

By default, the function returns data where each pollutant is in a separate column. However, it is possible to return the data in a *tidy* format (column for pollutant name, column for value) by using the option `to_narrow`:

```{r}
#| eval: false
my1 <- importUKAQ("my1", year = 2018, to_narrow = TRUE)
```

It is also possible to return information on whether the data have been ratified or not using the option `ratified` (`FALSE` by default). So, add the option `ratified = TRUE` if you want this information.

### Annual and other statistics

By default, all the functions above return hourly data. However, often there is a need to return data such as annual means of a long period of time. `importUKAQ()` (but not `importImperial()`) can return data for averaging times: annual, monthly, daily and for SO~2~ 15-minute. The annual and monthly data also provide valuable information on data capture rates. The averaging statistic is selected with the `data_type` option. The values `data_type` can take include:

-   **"hourly"** This is the default and specific site(s) must be provided.
-   **"daily"** Daily means returned and specific site(s) must be provided. Note that in the case of PM~10~ and PM~2.5~ daily measurements can be available from those derived from hourly measurements (using instruments such as TEOM, BAM and FIDAS) and from daily gravimetric measurements such as from a Partisol. In the returned data the gravimetric daily measurements are shown as `gr_pm10` and `gr_pm2.5`, respectively.
-   **"monthly"** Monthly means returned. No site code is needed because all data for a particular year are returned. Data capture statistics are also given.
-   **"annual"** Annual means returned. No site code is needed because all data for a particular year are returned. Data capture statistics are also given.
-   **"15_min"** 15-minute SO~2~ concentrations returned for a specific site(s).
-   **"8_hour"** Rolling 8-hour concentrations returned for a specific site(s) for O~3~ and CO.
-   **"24_hour"** Rolling 24-hour concentrations returned for a specific site(s) for PM~10~ and PM~2.5~.
-   **"daily_max_8"** Maximum daily rolling 8-hour maximum for O~3~ and CO.
-   **"daqi"** Daily Air Quality Index (DAQI). See [here](https://uk-air.defra.gov.uk/air-pollution/daqi?view=more-info&pollutant=ozone#pollutant) for more details of how the index is defined.

Note that for annual and monthly statistics all network data is returned and the `site` option has no effect.

As an example, to import 5 years of annual mean data from the AURN:

```{r}
#| eval: false
uk_annual <- importUKAQ(year = 2016:2020, data_type = "annual", source = "AURN")
```

By default, this will return data in "wide" format with a pollutant and its data capture rate in separate columns. Often it is more useful to have "narrow" format data, which is possible to select with the `to_narrow` option. Furthermore, it is also possible to return site meta data (site type, latitude and longitude) at the same time.

Below is an example of obtaining annual mean data for 2020.

```{r}
#| label: importAnnual
uk_2020 <- importUKAQ(
  year = 2020,
  source = "aurn",
  data_type = "annual",
  meta = TRUE,
  to_narrow = TRUE
)

uk_2020
```

The pollutants returned include:

```{r}
unique(uk_2020$species)
```

Now it is easy for example, to select annual mean data from 2020 for NO~2~ with a data capture rate of at least 80%:

```{r}
#| label: filterNO2
filter(uk_2020, species == "no2", data_capture >= 0.8)
```

For the AURN, AQE, SAQN, WAQN and NI networks, it is also possible to return the DAQI (Daily Air Quality Index) by pollutant to save deriving it.

```{r}
#| label: daqi
daqi_2020 <- importUKAQ(
  year = 2020,
  source = "aurn",
  data_type = "daqi",
  meta = TRUE
)

daqi_2020
```
